Beyond May: Complexity-stability relationships \\ in disordered dynamical systems

### 2023-04-30 papers/beyond-may.tex
housekeeping
edit draft

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}

\end{abstract}
\maketitle

\section*{Introduction}

\section*{Results}

@@

\paragraph*{\bf Random interactions: general reults.}
\input{random-general.tex}
\paragraph*{\bf Random interactions: power laws case.}
\input{random-power.tex}

\section{Conclusion}
\input{conclusion.tex}

\begin{acknowledgments}
We gratefully acknowledge multiple discussions with Ian Hatton, Matthieu Barbier and Ada Altieri on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

\clearpage

\bibliography{general-stability}
\bibliographystyle{unsrt}

>>>>

\date{\today}

\begin{abstract}
\input{abstract.tex}
\end{abstract}
\maketitle

\paragraph*{\bf Introduction.}
\input{intro.tex}

@@

\paragraph*{\bf Random interactions: stability condition.}
\input{stability.tex}

\paragraph*{\bf Random interactions: equilibrium distribution.}
\input{equilibrium.tex}

\paragraph*{\bf Conclusion.}
\input{conclusion.tex}

\medskip

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for many productive discussions on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

\medskip

\bibliography{beyond-may}
\bibliographystyle{unsrt}

### 2023-04-30 papers/conclusion.tex
edit draft

The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle.

We reconsidered the relationship between diversity and stability in complex systems, a classic problem since May's 1973's puzzling prediction that "complexity begets instability".

Considering dynamical models more general than the classical Lotka-Volterra model, we found that the relation can in fact go both ways:  depending on the relative convexity of self- and cross-interactions,

>>>>

The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and in other fields. 
In this letter, we have showed that the condition $\sigma\sqrt{N}< \mu_s - \mu$ (popularized by May's slogan ``complexity begets instability") does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that generalized Lotka-Volterra model, often cited in support of May's prediction, corresponds to a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where the self-dynamics is more strongly regulated than the cross-interactions ($\alpha < \beta$, which might be more appropriate for ecological modeling \cite{Hatton2023}), the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

### 2023-04-30 papers/homogeneous.tex
edit draft

Under what condition does \eqref{dynamics} admit a linearly stable equilibrium? 
We begin by considering the simple case where all self-interactions have the same strength $a_{ii} = a_{\textrm{s}}$, and similarly for cross-interactions $a_{ij} = a_{\textrm{c}}$ ($i\neq j$).
Under these assumptions, a straightforward computation shows that a homogeneous equilibrium $\mathbf x_*$ will be linearly stable if  
\begin{equation}\label{homogeneous-general}
   \frac{f'(x^*)}{f(x^*)} - \frac{g'(x^*)}{g(x^*)} + (a_{\textrm{s}} - a_{\textrm{c}})\,\frac{g(x^*)h'(x^*)}{f(x^*)} < 0.
\end{equation}
Thus, stability of an equilibrium depends on the relative strength of self- and cross-interactions ($a_{\textrm{s}} - a_{\textrm{c}}$), but also on the relative growth rate of the production and response functions near that equilibrium ($f'_*/f_* - g'_*/g_*$).
With power laws, \eqref{homogeneous-general} evaluates to 
\begin{equation}
    (\alpha - \beta)(N-1) < \gamma(a_{\textrm{s}}/a_{\textrm{c}}- 1) - (\alpha - \beta)(a_{\textrm{s}}/a_{\textrm{c}}),
\end{equation}
leading to three different regimes:
\begin{itemize}
    \item If $\alpha = \beta$, stability requires $a_{\textrm{s}} > a_{\textrm{c}}$, i.e.
    self-interactions must be stronger than cross-interactions.
    This is the usual conclusion drawn from the Lotka-Volterra model.
    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
    We can call this "May" behavior.
    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
    This is "anti-May" behavior.
\end{itemize}

>>>>

Under what condition does \eqref{dynamics} admit a (linearly) stable equilibrium? 
We begin with the simple case where all self-interactions have the same strength $A_{ii} = \mu_s$, and similarly for cross-interactions $A_{ij} = \mu$ ($i\neq j$).
Defining 
\begin{equation}
    \psi(x) \equiv (\mu - \mu_s) -  \left(\frac{f'(x)}{f(x)} - \frac{g'(x)}{g(x)}\right)\frac{f(x)}{g(x)h'(x)},
    \label{eq: psi}
\end{equation}
we find that stability of the homogeneous equilibrium $x_i^* = x^*$ requires $\psi(x^*) > 0$. This condition involves the relative strength of diagonal and off-diagonal interactions ($\mu_s - \mu$), but also on the relative growth rate of the production and response functions near the equilibrium ($f'_*/f_* - g'_*/g_*$).

With power laws, the condition $\psi(x^*)<0$ evaluates to 
\begin{equation}
    (\alpha - \beta)(N-1) < \gamma(\mu_s/\mu- 1) - (\alpha - \beta)(\mu_s/\mu),
\end{equation}
leading to three different regimes:
\begin{itemize}
    \item If $\alpha = \beta$, stability requires $\mu_s > \mu$, i.e.
    self-interactions must be stronger than cross-interactions.
    This is the usual conclusion drawn from the Lotka-Volterra model.
    \item If $\alpha > \beta$, stability places an upper bound on $N$: the more complex the system, the less likely to be stable.
    We can call this ``May" behavior.
    \item If $\alpha < \beta$, stability places an lower bound on $N$: the more complex the system, the more likely to be stable.
    This is ``anti-May" behavior.
\end{itemize}

### 2023-04-30 papers/intro.tex
edit draft

Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also outside biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology by studying in the first place? Empirically, the opposite relationship between complexity and stability seems to hold across ecosystems: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and  observation is at the center of the longlasting "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.

May's argument can be summarized as follows.
A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $\mathbf x^*$.
Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions (with zero 
mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale, 
then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, to correlated interactions, see e.g. \cite{allesina2015stability}.)

In ecology, many authors have sought to ease the tension between May's 
prediction and empirical observation by invoking effects not captured by 
dynamical systems with random 
coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: 
random dynamical systems do not necessarily imply that 
stability decreases with dimensionality and interaction 
strength---the opposite behavior is also possible, 
without the need for special or additional structure.

>>>>

Few mathematical arguments have influenced biological thinking like May's prediction that large, complex ecosystems cannot be stable~\cite{May1972}.
It is, on the face of it, a perplexing conclusion.
On the one hand, May's mathematical argument is simple and seemingly model-free, suggesting universal applicability, also beyond biology \cite{Haldane2011, Moran2019}.
On the other hand, it is clear that at least some large, complex systems are stable---else which regularities would biology by studying in the first place? 
Empirically, the opposite relationship between complexity and stability seems to hold: species-rich, strongly-coupled communities such as rainforests tend to be stable over time, while sparser ones, for instance arctic communities, often exhibit large fluctuations, exctinctions, and invasions~\cite{Hutchinson1959,Odum1959,MacArthur1955}. 
The tension between May's theoretical argument and  observation is at the center of the longstanding "diversity-stability debate" in ecology \cite{McCann2000, Loreau2022}.

May's argument can be summarized as follows.
A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $\mathbf x^*$.
Near that equilibrium $\mathbf x = \mathbf x^* + \delta \mathbf x$, the dynamics of the system is described by linear equations $d(\delta \mathbf x)/dt = A (\delta \mathbf x)$, and the stability of these equations requires that all eigenvalues of $A$ have negative real part.
But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions (with zero 
mean and variance $\sigma^2$), and $-I$ corresponds to stabilizing self-interactions on some natural timescale, 
then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$.
This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability".
(This argument generalizes to $\langle B_{ij}\rangle \neq 0$, incomplete connectivity, or correlated interactions, see e.g. \cite{allesina2015stability}.)

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients~\cite{McCann2000,Chesson2000,Mougi2012,Rohr2014,Barabas2017,Grilli2017}. 
In this letter, we use recent results in the physics of disordered systems \cite{Ahmadian2015, Roy2019} to show that May's argument itself is incomplete: random dynamical systems do not necessarily imply that stability decreases with dimensionality and interaction strength---the opposite behavior is also possible, without the need for special or additional structure.
For an in-depth discussion of ecological implications of our result, see \cite{Mazzarisi2023}.

### 2023-04-30 papers/model.tex
housekeeping
edit draft

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp.
$a_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$.
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.
This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
From a different perspective, it has been suggested recently that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

>>>>

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$. 
We all assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.

This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

The classic generalized Lotka-Volterra (GLV) competition model studied recently by Bunin and collaborators \cite{bunin2017ecological, biroli2018marginally} corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$.
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$.
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$~\cite{Brown2004}, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks.
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level~\cite{Hatton2015,Hatton2023}.
In a different direction, it has been recently suggested that predator-prey interactions can be modelled with a square root law ($g(x) \sim h(x) \sim x^{1/2}$)~\cite{Barbier2021,Mazzarisi2023}.

### 2023-04-30 papers/random-gen.tex
edit draft

To use this result, we decompose the interaction matrix as $a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Using rank-one perturbation theory, we can show that the $\mathbf{1}$ term does not impact stability properties and can be neglected \cite{Stone2018}.
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} thus takes the form $M + LJR$ with diagonal matrices
\begin{equation}
    M = f(\mathbf x^*)\psi(\mathbf x^*),\quad L = g(\mathbf x^*), \quad R = h'(\mathbf x^*),
\end{equation}
where $\psi(x)$ is the function defined by the LHS of \eqref{homogeneous-general}. \red{Assuming all $M_i < 0$}, the domain \eqref{eq: domain} containing the eigenvalues of $J^*$ first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires
\begin{equation}
    \sum_{i=1}^N \left(\frac{g(x_i^*)h'(x_i^*)}{f(x_i^*)\psi(x_i^*)}\right)^2
    < \frac{1}{\sigma^{2}}
    \label{eq: random-stability}
\end{equation}
This expression generalizes \eqref{homogeneous-general} to random interactions.

Straightforward results follow from condition \eqref{eq: random-stability}.
For GLV models, i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$, \eqref{eq: random-stability} is independent of the equilibria $x^*$ and we recover the linear stability condition
$\sigma\sqrt{N} < (\mu_s-\mu)$.
More generally, any model satisfying
\begin{equation}\label{eq: equiv class glv}
    \left(\cfrac{f'(x_i^*)}{f(x_i^*)} -
        \cfrac{g'(x_i^*)}{g(x_i^*)}\right)\Bigl /
        \cfrac{g(x_i^*)h'(x_i^*)}{f(x_i^*)} = c \ ,
\end{equation}
for some constant $c$ has the same simple linear stability condition as GLV models, namely $\sigma\sqrt{N} < (\mu_s-\mu) +c$.

When the dependence on $\mathbf{x}_*$ does not cancel out, it is not possible to obtain an explicit complexity-stability relationship. 
However, introducing the distribution of equilibrium $P(x^*)$ with corresponding equilibrium average $\langle\cdot\rangle$, we can write the stability condition as
\begin{equation}\label{eq: random-stability int}
    N\Big\langle\frac{gh'}{f\psi}\Big\rangle < \sigma^{-2} \ .
\end{equation}
We now discuss how $P(x_*)$ can be estimated from the statistics of interactions, focusing on the power-law case.

>>>>

To use this result, we decompose the interaction matrix as $a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Using rank-one perturbation theory, we can show that the $\mathbf{1}$ term does not impact stability properties and can be neglected \cite{Stone2018}.
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} thus takes the form $M + LJR$ with diagonal matrices
\begin{equation}
    M_i = f(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*).
\end{equation}
\red{Assuming all $M_i < 0$}, the domain \eqref{eq: domain} containing the eigenvalues of $J^*$ first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires
\begin{equation}
    \sum_{i=1}^N \left(\frac{g(x_i^*)h'(x_i^*)}{f(x_i^*)\psi(x_i^*)}\right)^2
    < \frac{1}{\sigma^{2}}
    \label{eq: random-stability}
\end{equation}
This expression generalizes \eqref{homogeneous-general} to random interactions.

For GLV models, i.e. $f(x_i)=x_i$, $g(x_i)=x_i$ and $h(x_i)=x_i$, \eqref{eq: random-stability} is independent of the equilibria $x^*$, and we recover the linear stability condition
$\sigma\sqrt{N} < (\mu_s-\mu)$.
More generally, any model satisfying
\begin{equation}\label{eq: equiv class glv}
    \left(\cfrac{f'(x_i^*)}{f(x_i^*)} -
        \cfrac{g'(x_i^*)}{g(x_i^*)}\right)\Bigl /
        \cfrac{g(x_i^*)h'(x_i^*)}{f(x_i^*)} = c \ ,
\end{equation}
for some constant $c$ has the same simple linear stability condition as GLV models, namely $\sigma\sqrt{N} < (\mu_s-\mu) +c$.

When the dependence on $\mathbf{x}_*$ does not cancel out, it is not possible to obtain an explicit complexity-stability relationship. 
However, introducing the distribution of equilibrium $P(x^*)$ and replacing sums with integrals in the large $N$ limit, we can write the stability condition as
\begin{equation}\label{eq: random-stability int}
    N\int dx^* frac{g(x^*)h'(x^*)}{f(x^*)\psi(x^*)} < \sigma^{-2} \ .
\end{equation}
We now discuss how $P(x_*)$ can be estimated from the statistics of interactions, focusing on the power-law case.
